Forecasting short-term hourly Emergency Departement arrivals

### 2021-12-09 paper/paper.Rmd
McCarthy discussion and minor correction up to GAMLSS section

Linear regression, ARIMA, and Naïve were used by @hertzum2017forecasting to investigate whether accurate hourly accident and emergency department patient arrivals and occupancy forecasts can be generated using calendar variables. Naïve method was there for the purpose of comparison. @hertzum2017forecasting study shows that patient arrivals variation is larger across the hours of the day than across the days of the week and the months of the year. In terms of hour of the day, patient arrivals peaked around noon. For days of the week, Monday is the busiest day while weekends are the quietest ones. July-August are the months with the highest numbers of patient arrivals, while January and February are the months with the lowest numbers. According to @hertzum2017forecasting, the regression and ARIMA models performed similarly in modelling patient arrivals, while ARIMA outperformed regression models in modelling accident and emergency department occupancy. However, after all, the models of occupancy were less accurate than those of arrivals. @hertzum2017forecasting mentioned that ARIMA was among the most accurate models in terms of forecasting accident and emergency department visits.
<!-- Another interesting point is that the accuracy of accident and emergency department forecasting models decrease with the increasing forecast interval. Lastly, the accuracy of the forecasting model may possibly be increased with additional information added to the model. -->

@@

@mccarthy2008challenge employed a Poisson log-linear regression model, including independent variables such as temporal factors (e.g., hour-of-day, day- of-week, type-of-day, season, and calendar year), patient characteristics (i.e., age, gender, insurance status, triage level, mode of arrival, and ambulance diversion status) and climatic factors (i.e., temperature and precipitation) to forecast patient demand for ED services. Hourly data of ED arrivals in the 1-year study period was deployed to forecast from 1 hour to 24 hours into the future. The authors presented the prediction intervals with $50\%$ and $90\%$ confidence levels for the number of hourly arrivals under the Poisson assumption. They showed that the most important predictor was hour of the day and AR(1).
<!-- !!!! IS: I don't understand what is meant by the next sentence !!!! -->
<!-- However, these findings are limited to the short number of observations (only one year of historical data). -->

@@

Data used in this study comprises counts of patients’ arrival times at one of the largest ED units in the UK between April 2014 and February 2019, extracted from the ED administrative database of the hospital. We aggregated the patients’ arrival times to obtain hourly arrivals, which is used for empirical evaluation in this study.

Figure \@ref(fig:hourly-plot-ridge) illustrates the distribution of arrivals for each hour of the day and the day of the week. Although the data is noisy, it reveals some systematic patterns.

@@

Figure \@ref(fig:hourly-plot-ridge) also highlights that there are many outliers for almost every hour of the day that may affect the accuracy of forecasting methods. Some of these outliers might be related to holidays and special events. It is also clear that arrivals are less volatile between mid-night and early morning.

@@

Based on this simple analysis, we should consider model that could take the following into account:

- Hour of day, day of week, and week of year seasonal patterns,
- Calendar events, such as holidays,
- Lags of calendar events to accommodate the potential changes in demand the next day after a holiday,
- Potentially changing level of data,
- Other events, such as football matches in the area.

We propose several forecasting models that account for the structures outlined above.

@@

## Multiple Regression
Regression is one of the most popular forecasting methods that uses explanatory variables to predict a variable of interest (in our case, the ED arrivals). The classical linear regression model is formulated as:

\begin{equation}
  {y}_t = \mathbf{x}_t' \boldsymbol{\beta} + \epsilon_t ,
(\#eq:linearRegression)
\end{equation}

where $\mathbf{x}_t$ is the vector of explanatory variables, $\boldsymbol{\beta}$ is the vector of parameters, $\epsilon_t$ is the error term, which is typically assumed to follow Normal distribution with zero mean and a fixed variance, and $t$ is the time index. However, in the context of healthcare and ED arrivals, the assumption of Normality is unrealistic, because the number of admitted patients is integer and non-negative. So the linear regression model should be substituted by some other model. One of the models that is frequently used in practice is the Poisson regression [see for example, @mccarthy2008challenge], which can be summarised as:

\begin{equation}
  \begin{aligned}
    & {y}_t \sim \mathrm{Poisson} \left( \lambda_t \right) \\
    & \log \lambda_t = \mathbf{x}_t' \boldsymbol{\beta}
  \end{aligned}
(\#eq:PoissonRegression)
\end{equation}

@@

The variables (1) - (3) allow modelling the seasonal patterns on the appropriate level of detail throughout the year, while (4) covers the changes in admittance due to calendar events. Finally (5) is needed in order to capture the potential phenomenon of change in admittance after the holiday (e.g. people might try not to go to hospital on Christmas eve and thus will go the next day). This model assumes that all these effects are deterministic and do not change over time, but the exponentiation in \@ref(eq:PoissonRegression) introduces an interaction effect between dummy variables, so that the 3pm on Monday in January will be different from 3pm on Monday in July, although the parameters for hour of day and day of week are fixed and do not change over time. We use `alm()` function from `greybox` package [@Svetunkov2021Greybox] for R [@RTeam2021] for the experiments and denote this model as "Poisson Regression".

## ETS - Exponential Smoothing model

@@

@de2011forecasting proposed a model to deal with time series exhibiting multiple complex seasonalities called "TBATS". It includes a Box-Cox Transformation, ARMA model for residuals and a trigonometric expression of seasonal terms. The latter one not only gives the model more flexibility to deal with fractional seasonality but also reduces the parameters of model when the frequencies of seasonalities are high. We fit a TBATS model using the `tbats()` function from the `forecast` package in R [@forecastpackage2020].

@@

@SvetunkovAdam2021 proposed a framework for dynamic models called **ADAM** - Augmented Dynamic Adaptive Model. This framework encompasses ARIMA [@Box1976], ETS [@Hyndman2008b] and regression, supporting multiple frequencies, non-normal distributions and intermittent demand [@Svetunkov2019a]. Based on this framework, we use Gamma distribution for ETS(M,N,M) model with frequencies 24 (hour of day) and 168 (hour of week), adding dummy variables for week of year, holidays and lagged holidays. This way we update the hour of day and day of week seasonal indices, keeping the week of year one fixed, thus reducing the number of estimated parameters. Given that the data exhibits randomly occurring zeroes, we use the direct probability model developed by @Svetunkov2019a to treat those values. This model can be formulated as a set of the following equations:

\begin{equation}
	\begin{aligned}
	  & y_t = o_t z_t \\
		& \log z_t = \log l_{t-1} + \log s_{1,t-24} + \log s_{2,t-168} + \mathbf{x}_t' \boldsymbol{\beta} + \log \left(1 + \epsilon_{t} \right) \\
		& \log l_{t} = \log l_{t-1} + \log( 1  + \alpha \epsilon_{t}) \\ 
		& \log s_{1,t} = \log s_{1,t-m} + \log( 1  + \gamma_1 \epsilon_{t}) \\
		& \log s_{2,t} = \log s_{2,t-m} + \log( 1  + \gamma_2 \epsilon_{t}) \\
		& o_t \sim \text{Bernoulli} \left(\mu_{a,t} \right) \\
		& a_t = l_{a,t-1} \left(1 + \epsilon_{a,t} \right) \\
		& l_{a,t} = l_{a,t-1}( 1  + \alpha_{a} \epsilon_{a,t}) \\
		& \mu_{a,t} = \min(l_{a,t-1}, 1)
	\end{aligned} ,
	(\#eq:ADAMModel)
\end{equation}
where $\alpha$, $\beta$, $\gamma_1$, $\gamma_2$ and $\alpha_a$ are the smoothing parameters, defining how adaptive the components of the model should be, $l_t$ is the level component for the demand sizes, $s_{1,t}$ and $s_{2,t}$ are the seasonal components, $\boldsymbol{\beta}$ is the vector of parameters for the explanatory variables, $o_t$ is the binary variable, which is equal to one, when demand occurs and to zero otherwise, $l_{a,t-1}$ is the level component for the occurrence part of the model, and $\left(1+\epsilon_t \right) \sim \Gamma(s^{-1}, s)$, where $s=\frac{1}{T} \sum_{t=1}^{T} e_{t}^2$ is the scale of the distribution. Finally, $a_t$ is an unobservable series, underlying the occurrence part of the model and $(1 + \epsilon_{a,t})$ is an unobservable error term for $a_t$. @Svetunkov2019a discuss how to estimate such model. We expect this model to perform on par with the Poisson regression, potentially outperforming it in some instances, due to the dynamic parts of the model (level and seasonal components). Although the data is integer-valued, we expect that Gamma distribution will be a good approximation for it. If integer-valued quantiles are needed, then rounding up can be done for them (see \@ref(quantiles_ceiling) for the explanation). This model is implemented in `adam()` function from `smooth` package [@Svetunkov2021Smooth] for R and is denoted in our experiment as "ADAM-iETSX".

## GAMLSS
If we assume that our predictive distribution follows a given parametric distribution, the forecasting task becomes ones of predicting the future values of that distribution's parameters. Generalised Additive Models for Location, Scale and Shape (GAMLSS) are distributional regression models where the parameters may be modelled as additive functions of explanatory variables. This provides a powerful and flexible framework for probabilistic forecasting, provided that a suitable distribution and additive structures can be found. In practice, this means experimenting with various distributions and evaluating their suitability using available training data.

>>>>

Linear regression, ARIMA, and Naïve were used by @hertzum2017forecasting to investigate whether accurate hourly accident and emergency department patient arrivals and occupancy forecasts can be generated using calendar variables. Naïve method was there for the purpose of comparison. @hertzum2017forecasting study found that patient arrival variation is larger across the hours of the day than across the days of the week and the months of the year. In terms of hour of the day, patient arrivals peaked around noon. For days of the week, Monday is the busiest day while weekends are the quietest ones. July-August are the months with the highest numbers of patient arrivals, while January and February are the months with the lowest numbers. According to @hertzum2017forecasting, the regression and ARIMA models performed similarly in modelling patient arrivals, while ARIMA outperformed regression models in modelling accident and emergency department occupancy. However, after all, the models of occupancy were less accurate than those of arrivals. @hertzum2017forecasting mentioned that ARIMA was among the most accurate models in terms of forecasting accident and emergency department visits.
<!-- Another interesting point is that the accuracy of accident and emergency department forecasting models decrease with the increasing forecast interval. Lastly, the accuracy of the forecasting model may possibly be increased with additional information added to the model. -->

@@

@mccarthy2008challenge employed a Poisson log-linear regression model, including independent variables such as temporal factors (e.g., hour-of-day, day- of-week, type-of-day, season, and calendar year), patient characteristics (i.e., age, gender, insurance status, triage level, mode of arrival, and ambulance diversion status) and climatic factors (i.e., temperature and precipitation) to model patient demand for ED services. The authors produced probabilistic predictions in the form of $50\%$ and $90\%$ prediction intervals for the number of hourly arrivals. Hourly data of ED arrivals in the 1-year study period was modelled and analysed, and it is suggested that the model could be used for forecasting, however, model evaluation was performed in-sample on only one year of data, so unclear how this approach would perform in a forecasting setting or compare to simpler approaches.
<!-- !!!! IS: I don't understand what is meant by the next sentence !!!! -->
<!-- However, these findings are limited to the short number of observations (only one year of historical data). -->

@@

Data used in this study comprises counts of patients’ arrival times at one of the largest ED units in the UK between April 2014 and February 2019, extracted from the ED administrative database of the hospital. We aggregated the patients’ arrival times to obtain hourly arrivals, which are used in this study. Figure \@ref(fig:hourly-plot-ridge) illustrates the distribution of arrivals for each hour of the day and the day of the week. Although the data is noisy, it reveals some systematic patterns.

@@

Figure \@ref(fig:hourly-plot-ridge) also highlights that there is significant skewness for almost every hour of the day that varies with time-of-day, and which should be accounted for in forecasting methods. Some skewness might be related to holidays and special events. It is also clear that arrivals are less volatile between mid-night and early morning.

@@

Based on this simple analysis and literature review, we should consider models that can take the following into account:
- Daily, weekly, and annual seasonalities,
- Long-term trend,
- Calendar events, such as holidays,
- Lags of calendar events to accommodate the potential changes in demand the next day after a holiday,
- Other events, such as sporting fixtures,
- Weather effects, such as temperature and precipitation.
We propose several forecasting models that account for the structures outlined above.

@@

## Poisson Regression
Regression is one of the most popular forecasting methods that uses explanatory variables to predict a variable of interest (in our case, the ED arrivals). The classical linear regression model is formulated as

\begin{equation}
  {y}_t = \mathbf{x}_t' \boldsymbol{\beta} + \epsilon_t ,
(\#eq:linearRegression)
\end{equation}
where $\mathbf{x}_t$ is the vector of explanatory variables, $\boldsymbol{\beta}$ is the vector of parameters, $\epsilon_t$ is the error term, which is typically assumed to follow Normal distribution with zero mean and a fixed variance, and $t$ is the time index. However, in the context of healthcare and ED arrivals, the assumption of Normality is unrealistic, because the number of admitted patients is integer and non-negative. So the linear regression model should be substituted by some other model. One of the models that is frequently used in practice is the Poisson regression [see for example, @mccarthy2008challenge], which can be summarised as

\begin{equation}
  \begin{aligned}
    & {y}_t \sim \mathrm{Poisson} \left( \lambda_t \right) \\
    & \log \lambda_t = \mathbf{x}_t' \boldsymbol{\beta}
  \end{aligned} \quad .
(\#eq:PoissonRegression)
\end{equation}

@@

The variables (1)--(3) allow modelling the seasonal patterns on the appropriate level of detail throughout the year, while (4) covers the changes in admittance due to calendar events. Finally (5) is needed in order to capture the potential phenomenon of change in admittance after the holiday (e.g. people might try not to go to hospital on Christmas eve and thus will go the next day). This model assumes that all these effects are deterministic and do not change over time, but the exponentiation in \@ref(eq:PoissonRegression) introduces an interaction effect between dummy variables, so that the 3pm on Monday in January will be different from 3pm on Monday in July, although the parameters for hour of day and day of week are fixed and do not change over time. We use `alm()` function from `greybox` package [@Svetunkov2021Greybox] for R [@RTeam2021] for the experiments and denote this model as "Poisson Regression".

## ETS - Exponential Smoothing

@@

@de2011forecasting proposed a model to deal with time series exhibiting multiple complex seasonalities called "TBATS". It includes a Box-Cox Transformation, ARMA model for residuals and a trigonometric expression of seasonal terms. The latter not only gives the model more flexibility to deal with fractional seasonality but also reduces the parameters of model when the frequencies of seasonalities are high. We fit a TBATS model using the `tbats()` function from the `forecast` package in R [@forecastpackage2020].

@@

@SvetunkovAdam2021 proposed a framework for dynamic models called the Augmented Dynamic Adaptive Model (ADAM). This framework encompasses ARIMA [@Box1976], ETS [@Hyndman2008b] and regression, supporting multiple frequencies, non-normal distributions and intermittent demand [@Svetunkov2019a]. Based on this framework, we use Gamma distribution for ETS(M,N,M) model with frequencies 24 (hour of day) and 168 (hour of week), adding dummy variables for week of year, holidays and lagged holidays. This way we update the hour of day and day of week seasonal indices, keeping the week of year one fixed, thus reducing the number of estimated parameters. Given that the data exhibits randomly occurring zeroes, we use the direct probability model developed by @Svetunkov2019a to treat those values. This model can be formulated as a set of the following equations:

\begin{equation}
	\begin{aligned}
	  & y_t = o_t z_t \\
		& \log z_t = \log l_{t-1} + \log s_{1,t-24} + \log s_{2,t-168} + \mathbf{x}_t' \boldsymbol{\beta} + \log \left(1 + \epsilon_{t} \right) \\
		& \log l_{t} = \log l_{t-1} + \log( 1  + \alpha \epsilon_{t}) \\ 
		& \log s_{1,t} = \log s_{1,t-m} + \log( 1  + \gamma_1 \epsilon_{t}) \\
		& \log s_{2,t} = \log s_{2,t-m} + \log( 1  + \gamma_2 \epsilon_{t}) \\
		& o_t \sim \text{Bernoulli} \left(\mu_{a,t} \right) \\
		& a_t = l_{a,t-1} \left(1 + \epsilon_{a,t} \right) \\
		& l_{a,t} = l_{a,t-1}( 1  + \alpha_{a} \epsilon_{a,t}) \\
		& \mu_{a,t} = \min(l_{a,t-1}, 1)
	\end{aligned} ,
	(\#eq:ADAMModel)
\end{equation}
where $\alpha$, $\beta$, $\gamma_1$, $\gamma_2$ and $\alpha_a$ are the smoothing parameters, defining how adaptive the components of the model should be, $l_t$ is the level component for the demand sizes, $s_{1,t}$ and $s_{2,t}$ are the seasonal components, $\boldsymbol{\beta}$ is the vector of parameters for the explanatory variables, $o_t$ is the binary variable, which is equal to one, when demand occurs and to zero otherwise, $l_{a,t-1}$ is the level component for the occurrence part of the model, and $\left(1+\epsilon_t \right) \sim \Gamma(s^{-1}, s)$, where $s=\frac{1}{T} \sum_{t=1}^{T} e_{t}^2$ is the scale of the distribution. Finally, $a_t$ is an unobservable series, underlying the occurrence part of the model and $(1 + \epsilon_{a,t})$ is an unobservable error term for $a_t$. @Svetunkov2019a discuss how to estimate such a model. We expect this model to perform on par with the Poisson regression, potentially outperforming it in some instances, due to the dynamic parts of the model (level and seasonal components). Although the data is integer-valued, we expect that Gamma distribution will be a good approximation for it. If integer-valued quantiles are needed, then rounding up can be done for them (see \@ref(quantiles_ceiling) for the explanation). This model is implemented in `adam()` function from `smooth` package [@Svetunkov2021Smooth] for R and is denoted in our experiment as "ADAM-iETSX".

## GAMLSS
If we assume that our predictive distribution follows a given parametric distribution, as in Poisson regression discussed above, the forecasting task becomes ones of predicting the future values of that distribution's parameters. Generalised Additive Models for Location, Scale and Shape (GAMLSS) are distributional regression models where the parameters are modelled as additive functions of explanatory variables. This provides a powerful and flexible framework for probabilistic forecasting, provided that a suitable distribution and additive model structures can be found. In practice, this means employing expert judgement and experimenting with various distributions and evaluating their suitability using available training data.
